---
title: "Storm and Weather Events Data Analysis (collected since 1970 - 2011)"
output: 
html_document:
keep_md: true
---
###Reproducible Research: Peer Assessment 2       

##Synopsis:    
Storm and Weather events causes harmful for population health and damage property which impacts country's economic conditions. U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database is the 
collection of data from various parts of the country to study/analyze the cuases of more health and economic consequences to take preventive actions.  Results of this analysis address the following questions:  
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?  

##Processing Data     

The analysis was performed on [Storm Events Database]("http://www.ncdc.noaa.gov/stormevents/ftp.jsp"), provided by [National Climatic Data Center]("http://www.ncdc.noaa.gov/"). The data is from a comma-separated-value file available [here]("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"). There is also some documentation of the data available [here]("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf").  

Downloading data file
```{r 'init', echo=TRUE}
#setting local working directory
setwd("C:/Data/devtools/Git/RepData_PeerAssessment2")
library(knitr)
library(ggplot2)
#suppressMessages to suppress warning/ messages
suppressMessages(library(dplyr))
#setting working directory for knit
opts_knit$set(base.dir = "C:/Data/devtools/Git/RepData_PeerAssessment2")
stdata <- NULL
```
```{r 'downloaddatafile', echo=TRUE}
#Checking for file in current directory
if(!file.exists("PA2_StormData.bz2"))
{
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",destfile = "PA2_StormData.bz2",mode = "wb")
}

```  
Reading and checking data from file  
```{r 'readingdata', echo=TRUE, cache=TRUE}
#reading data from csv file
stdata <- read.csv(bzfile("PA2_StormData.bz2"))
#getting rows and columns count
colnms <- names(stdata)
#rows & columns
rws <- nrow(stdata); cls <- ncol(stdata)

```

Data from file:    
. Numer of rows **`r rws`**   
. Number of columns **`r cls`**  

Filtering required columns from dataframe for analysis. Ploting histogram to understand the data available for each year in U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database (from 1970-2011). 

```{r global_opts, include=FALSE}
knitr::opts_chunk$set(fig.width=20, fig.height=15, fig.path='figs/', warning=FALSE, message=FALSE)
```

```{r 'dataforanalysis', echo=TRUE}
#getting required data for analysis
prcdata <- stdata
names(prcdata) <- toupper(names(prcdata))
#getting required columns
prcdata <- prcdata[,c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
#formatting date
prcdata <- mutate(prcdata, BGN_DATE = as.Date(as.character(prcdata$BGN_DATE), "%m/%d/%Y"))
#Starting year
minDate <- min(prcdata$BGN_DATE)
maxDate <- max(prcdata$BGN_DATE)
#adding year column
prcdata$YEAR <- as.integer(format(prcdata$BGN_DATE, "%Y"))
hist(prcdata$YEAR, breaks = 45, main="Number of events recorded per year", xlab="Year", ylab="Number of events")
```


Histogram results supporting the statement **The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.**

Considerable measurements being collected from 1970 to 2011 by NOAA for major storms  and weather events.  
Event type (EVTYPE) values should be edited/updated with proper charecter sequences and trailing spaces to get proper counts and labels from the data.  

```{r 'harmful_events', echo=TRUE}
#filtering data from 1970 to 2011
strmdata <- filter(prcdata, YEAR >= 1970)
#converting to lower case
evntlbls <- toupper(strmdata$EVTYPE)
## Replace all punct. characters with a space
evntlbls <- gsub("(^[[:space:]]+|[[:space:]]+$)", "", evntlbls)
evntlbls <- gsub("[[:blank:][:punct:]+]", " ", evntlbls)
evntlbls <- gsub("^thunderstorm wind[:alnum:] | ^tstm wind[:alnum:]", "thunderstorm wind", evntlbls)
#updating data with updated labels
strmdata$EVTYPE <- evntlbls
#unique(strmdata$EVTYPE)
```  

##Data Analysis  

Subsetting wheather events which causes most harmful to population health and greatest economic consequences events from the data.  

##Major Weather Eevents harmful to Population Health##   
```{r 'HealthData', echo=TRUE}
#Getting harmful events data from dataframe
hdata <- filter(strmdata,strmdata$FATALITIES > 0 | strmdata$INJURIES > 0)
#harmful data rows count
nrow(hdata)
```

###Health Data Anasysis  

```{r echo=TRUE}
#Fatalities events counts
fatcounts <- aggregate(FATALITIES ~ EVTYPE,data=hdata,FUN=sum)
#InjuryEvents by aggregation
injcounts <- aggregate(INJURIES ~ EVTYPE,data=hdata,FUN=sum)
#Top ten records for FATALITIES and INJURIES
fatTop10 <- head(fatcounts[order(fatcounts$FATALITIES, decreasing = T), ], 10)
injTop10 <- head(injcounts[order(injcounts$INJURIES, decreasing = T), ], 10)
# Updating column names
colnames(fatTop10) <- c("Event", "Fatalities")
colnames(injTop10) <- c("Event", "Injuries")
```
##Results  
###Health Data Top 10 records   
. Fatal Events    
. Injury Events    


```{r 'healthdata_results', echo=TRUE}
fatTop10
injTop10
```

###Health Data plots  

```{r 'healthdata_plot', echo=TRUE}
par(mfrow = c(1, 2), mar = c(14, 6, 4, 3), mgp = c(2, 1, 0), cex = 0.9)

ylim <- c(0, 1.1*max(fatTop10$Fatalities))

fatalPlot <- barplot(fatTop10$Fatalities, names.arg = fatTop10$Event, main = 'Top 10 events for fatalities', ylab = 'Number of fatalities', ylim = ylim)
text(x = fatalPlot, y = fatTop10$Fatalities, label = round(fatTop10$Fatalities, 0), pos = 3, cex = 0.8)
ylim <- c(0, 1.1*max(injTop10$Injuries))

injuryPlot <- barplot(injTop10$Injuries, names.arg = injTop10$Event, main = 'Top 10 events for injuries', ylab = 'Number of injuries', ylim = ylim)
text(x = fatalPlot, y = injTop10$Injuries, label = round(injTop10$Injuries, 0), pos = 3, cex = 0.8)

```

##Economic Data Analysis  

```{r 'EconomicData', echo=TRUE}
#Economic consequence events data
edata <- filter(strmdata, strmdata$PROPDMG > 0 | strmdata$CROPDMG > 0)
#economic data rows count
nrow(edata)
```

```{r 'conversionFunction', echo=TRUE}
#Function to convert damage amount unit:
# h -> hundred, k -> thousand, m -> million, b -> billion
convertCurrUnit <- function(e) 
{
        if (e %in% c('h', 'H')){
                return(2)
        } else if (e %in% c('k', 'K')) {
                return(3)
        } else if (e %in% c('m', 'M')) {
                return(6)
        } else if (e %in% c('b', 'B')) {
                return(9)
        } else if (!is.na(as.numeric(e))) {# if a digit
                return(as.numeric(e))
        } else if (e %in% c('', '-', '?', '+')) {
                return(0)
        } else {
                stop("Not valid.")
        }
}
```

###Calculating Property and Corp damage dxpenses  

```{r 'economiccons', echo=TRUE}
#Getting property damage
edata$PROPDMG <- edata$PROPDMG * (10 ** sapply(edata$PROPDMGEXP, FUN=convertCurrUnit))
#Getting corp damage 
edata$CROPDMG <- edata$CROPDMG * (10 ** sapply(edata$CROPDMGEXP, FUN=convertCurrUnit))
```


```{r 'mosteconomicexpenses', echo=TRUE}
# Fatal events
prcounts <- aggregate(PROPDMG ~ EVTYPE,data=edata,FUN=sum)
crcounts <- aggregate(CROPDMG ~ EVTYPE,data=edata,FUN=sum)
# Events caused most economic expenses
prevntTop10 <- head(prcounts[order(prcounts$PROPDMG, decreasing = T), ], 10)
crevntTop10 <- head(crcounts[order(crcounts$CROPDMG, decreasing = T), ], 10)

# Updating column names
colnames(prevntTop10) <- c("Event", "propDMG")
colnames(crevntTop10) <- c("Event", "cropDMG")
```  
###Reselts  
###Economic Data Top 10 records    
. Property damage    
. Corp damage  

```{r 'economicdata_results', echo=TRUE}
prevntTop10
crevntTop10 
```  
###Economic data plots  

```{r echo=TRUE}
par(mfrow = c(1, 2), mar = c(12, 5, 3, 2), mgp = c(3, 1, 0), cex = 0.8, las = 3)

prdmgplot <- barplot((prevntTop10$propDMG/1000000000), names.arg = prevntTop10$Event, main = 'Top 10 events for fatalities', ylab = 'Number of fatalities (Billions)', log="y")
crdmgplot <- barplot((crevntTop10$cropDMG/1000000000), names.arg = crevntTop10$Event, main = 'Top 10 events for injuries', ylab = 'Number of injuries (Billions)', log="y")
```  

*Execute below script in commandline (or R console) to generate plot images and place them in './figure' folder  
knit2html("PA2_template.Rmd", "PA2_template.html")*  

